---
title: R Review Project
author: Kiley Huffman
date: "2024-11-14"
output: html_notebook
---
# Part 2: SKCM Analysis

In this project, you will conduct multi-omic analyses to explore the following research question:

What are the differences between metastatic and non-metastatic SKCM across the epigenome and do these have any effect on the transcriptome?

Exploration of Methylation Patterns and Effect on Transcription

# Set-Up
```{r}
# Set up knit statement
 knitr::opts_knit$set(root.dir = normalizePath("/home1/kehuffma/490_cluster/r_review"))
```

```{r}
# Download the Clinical Data

if (!require("TCGAbiolinks")) BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)

clin_query <- GDCquery(project = "TCGA-SKCM",
                      data.category = "Clinical",
                      data.type = "Clinical Supplement",
                      data.format = 'BCR Biotab')

GDCdownload(clin_query)

clinical.BCRtab.all <- GDCprepare(clin_query)
clinic <- clinical.BCRtab.all$clinical_patient_skcm[-c(1,2),]
colnames(clinic)[ colnames(clinic) == "bcr_patient_barcode" ] <-"Tumor_Sample_Barcode"
```

```{r}

# Create CSV for Clinical Data
write.csv(clinic, "/home1/kehuffma/490_cluster/r_review/skcm_clinical_data.csv",
          row.names = FALSE)
clinic <- read.csv("/home1/kehuffma/490_cluster/r_review/skcm_clinical_data.csv")
```

```{r}
# Download RNA data

if (!require("SummarizedExperiment")) BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)

skcm_query <- GDCquery(project ="TCGA-SKCM",
                      data.category = "Transcriptome Profiling",
                      data.type = "Gene Expression Quantification",
                      workflow.type = "STAR - Counts")

GDCdownload(skcm_query)
skcm_se <- GDCprepare(skcm_query)
```
```{r}

# Preprocess RNA data
skcm_clinical <- colData(skcm_se)[, ]
skcm_clinical <- as.data.frame(colData(skcm_se))
skcm_clinical <- skcm_clinical[, !(colnames(skcm_clinical) %in% c("treatments", "primary_site", "disease_type"))]
skcm_genes <- as.data.frame(elementMetadata(rowRanges(skcm_se)))
skcm_counts <- as.data.frame(skcm_se@assays@data$unstranded)

colnames(skcm_counts) <- skcm_clinical$patient
rownames(skcm_counts) <- skcm_genes$gene_id

# Mask out definition that are not of the type Metastatic/Nonmetastatic
tumor_mask <- skcm_clinical$definition != "Solid Tissue Normal" & 
              skcm_clinical$definition != "Additional Metastatic"
skcm_clinical <- skcm_clinical[tumor_mask, ]
skcm_clinical <- skcm_clinical[!is.na(skcm_clinical$definition),]
unique(skcm_clinical$definition)

# Mask Out NA (Not Reported) values
valid_vital <- !is.na(skcm_clinical$vital_status) & skcm_clinical$vital_status != "Not Reported"
```
```{r}
skcm_clinical <- skcm_clinical[valid_vital, ]
```

# 1. Difference in survival between metastatic and non-metastatic patients (KM plot):
```{r}
# Download packages if needed
if (!require(survival)) {install.packages("survival")}

if (!require(survminer)) {install.packages("survminer")}

if (!require(ggplot2)) {install.packages("ggplot2")}

library(survival)
library(survminer)
library(ggplot2)
library(dplyr) 

```

```{r}
# Prepare data
# Categorize Metastatic Status
skcm_clinical$metastatic_status <- ifelse(
skcm_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic")
skcm_clinical$metastatic_status <- factor(skcm_clinical$metastatic_status)

# Generate Survival Time Column
skcm_clinical$survival_time <- ifelse(is.na(skcm_clinical$days_to_death),
                                       skcm_clinical$days_to_last_follow_up,
                                       skcm_clinical$days_to_death)
skcm_clinical$survival_time <- as.numeric(skcm_clinical$survival_time)


skcm_clinical$death_event <-  ifelse(skcm_clinical$vital_status == "Dead", TRUE, FALSE)

```

```{r}

#KM plot
survival_object <- Surv(time = skcm_clinical$survival_time,
                        event = skcm_clinical$death_event)
fit_object <- survfit(survival_object ~ metastatic_status, data = skcm_clinical)


survplot <- ggsurvplot(fit_object,
                       pval=TRUE,
                       ggtheme = theme(plot.margin = unit(c(1,1,1,1), "cm")),
                       legend = 'right')

Plot1_KM <- survplot$plot + theme_bw() + theme(axis.title = element_text(size=20), 
                                              axis.text = element_text(size=16),
                                              legend.title = element_text(size=14),
                                              legend.text = element_text(size=12))
Plot1_KM

ggsave("/home1/kehuffma/490_cluster/r_review/Plot1_KM.pdf")

```

# 2. Differential expression between non-metastatic and metastatic patients controlling for treatment effects, race, gender, and vital status (DESeq2 + Volcano plot)
# a. Treatments must include radiation, chemotherapy, immunotherapy, molecular therapy, vaccine
# b. If you run this on CARC, it may take up to 1-2 hours

```{r}

# Get rad and drug data 
rad <- clinical.BCRtab.all$clinical_radiation_skcm [-c (1,2),]
drug <- clinical.BCRtab.all$clinical_drug_skcm [-c (1,2),]
drug$patient_id <- toupper(drug$bcr_patient_barcode)
rad$patient_id <- toupper(rad$bcr_patient_barcode)

unique(drug$pharmaceutical_therapy_type)

# Process Data for Categorization
drug$pharmaceutical_therapy_type <- gsub("Targeted Molecular therapy", "Molecular", drug$pharmaceutical_therapy_type)

treatment_types <- c("Chemotherapy", "Immunotherapy", "Molecular", "Vaccine")
for (treatment in treatment_types) {
    drug[[treatment]] <- ifelse(
        drug$pharmaceutical_therapy_type == treatment,
        1,
        0
    )
}

rad$Radiation_Therapy <- 1

# Match processed data with patient_id
drug_treatments <- drug[, c("patient_id", treatment_types)]
rad_treatments <- rad[, c("patient_id", "Radiation_Therapy")]

```

```{r}
# Aggregate Drug Treatments
drug_treatments <- drug %>%
  group_by(patient_id) %>%
  summarize(
    Chemotherapy = max(Chemotherapy, na.rm = TRUE),
    Immunotherapy = max(Immunotherapy, na.rm = TRUE),
    Molecular = max(Molecular, na.rm = TRUE),
    Vaccine = max(Vaccine, na.rm = TRUE)
  )

# Aggregate Radiation Treatments
rad_treatments <- rad %>%
  group_by(patient_id) %>%
  summarize(
    Radiation_Therapy = max(Radiation_Therapy, na.rm = TRUE)
  )

# Merge Treatment data by patient_id
treatment_data <- full_join(drug_treatments, rad_treatments, by = "patient_id")

# Replace NA values with 0
treatment_data[is.na(treatment_data)] <- 0

skcm_clinical$patient_id <- toupper(skcm_clinical$patient)

# Merge clinical data with treatment data
skcm_clinical <- left_join(skcm_clinical, treatment_data, by = "patient_id")

# Create new columns in skcm_clinical
treatment_columns <- c("Radiation_Therapy", "Chemotherapy", "Immunotherapy", "Molecular", "Vaccine")

skcm_clinical[, treatment_columns][is.na(skcm_clinical[, treatment_columns])] <- 0

for (treatment in treatment_columns) {
  skcm_clinical[[treatment]] <- factor(skcm_clinical[[treatment]])
}
```

```{r}
# Categorize Metastatic Status
skcm_clinical$metastatic_status <- ifelse(
  skcm_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic")
skcm_clinical$metastatic_status <- factor(skcm_clinical$metastatic_status)

# Mask out NAs and as.factor the covariants
skcm_clinical$race <- factor(skcm_clinical$race)
race_mask <- skcm_clinical$race != "not reported"
skcm_clinical <- skcm_clinical[race_mask,]
skcm_clinical$gender <- factor(skcm_clinical$gender)
skcm_clinical$vital_status <- factor(skcm_clinical$vital_status)
vital_mask <- skcm_clinical$vital_status != "Not Reported"
skcm_clinical <- skcm_clinical[vital_mask,]
```

```{r}
# Filter out duplicated patient ids
skcm_clinical$patient[duplicated(skcm_clinical$patient)]
skcm_clinical_unique <- skcm_clinical[!duplicated(skcm_clinical$patient), ]

# Matching rownames of clinical with columns of counts
clinical_samples <- skcm_clinical_unique$patient
counts_samples <- colnames(skcm_counts)
clinical_samples <- toupper(clinical_samples)
counts_samples <- toupper(counts_samples)
common_samples <- intersect(colnames(skcm_counts), skcm_clinical_unique$patient)
skcm_clinical_sub <- skcm_clinical[skcm_clinical_unique$patient %in% common_samples, ]
skcm_clinical_sub <- skcm_clinical_sub[match(common_samples, skcm_clinical_sub$patient), ]
skcm_counts_sub <- skcm_counts[, common_samples]

rownames(skcm_clinical_sub) <- skcm_clinical_unique$patient

gene_totals <- rowSums(skcm_counts_sub)

# Filter genes with total expression >= 1000
expressed_genes <- gene_totals >= 20
skcm_counts_filtered <- skcm_counts_sub[expressed_genes, ]

# Conduct DE
design_formula <- as.formula(paste("~", paste(
  c("Radiation_Therapy", "Chemotherapy", "Immunotherapy", "Molecular", "Vaccine", 
    "race", "gender", "vital_status", "metastatic_status"), collapse = " + ")))
```

```{r}
# Install DESeq2 if needed
if (!require("DESeq2", quietly = TRUE))
BiocManager::install("DESeq2")
library(DESeq2)

# Perforrm DESeq2 analysis
dds <- DESeqDataSetFromMatrix(
  countData = skcm_counts_filtered,
  colData = skcm_clinical_sub,
  design = design_formula
)

dds <- DESeq(dds)
```
```{r}

# Filter the results for NA
res <- results(dds, format = "DataFrame", contrast = c("metastatic_status", "Non-Metastatic" , "Metastatic"))
res <- res[!is.na(res$padj), ]
res_filtered <- res[
  res$padj < 0.05 & abs(res$log2FoldChange) > 1, 
]
res_df <- as.data.frame(res_filtered)

res_df$gene_id <- rownames(res_df)

# Adding gene_name and -log10(padj) to result dataframe
res_df <- merge(
  res_df, 
  skcm_genes[, c("gene_id", "gene_name")], 
  by = "gene_id", 
  all.x = TRUE
)

res_df[, "-log10(padj)"] <- -log10(res_df$padj)

```

```{r}

# Install Enhanced Volcano package if needed
if (!requireNamespace("EnhancedVolcano", quietly = TRUE))
    BiocManager::install("EnhancedVolcano")
library(EnhancedVolcano)

# Create volcano plot
EnhancedVolcano(res_df,
    lab = res_df$gene_name,  
    x = 'log2FoldChange',
    y = 'padj',
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10]~ 'adjusted p-value'),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    title = 'Volcano Plot of Differential Expression',
    subtitle = 'Thresholds: padj < 0.05 & |log2FC| > 1',
    caption = 'Source: DESeq2 Analysis'
)

ggsave("/home1/kehuffma/490_cluster/r_review/Plot2_Volcano.pdf")

write.csv(res_df, 'res_df.csv')

```

# 3. Methylation differences between metastatic and non-metastatic patients:
```{r}
# Download packages if needed
if (!require("sesameData", quietly = TRUE))
BiocManager::install("sesameData")

if (!require("sesame", quietly = TRUE))
BiocManager::install("sesame")

if (!require("limma", quietly = TRUE))
BiocManager::install("limma")
```


```{r}
# Load necessary packages
library(TCGAbiolinks)
library(sesame)
library(sesameData)
library(limma)

```

```{r}
# Change knit statement before downloading methylation data
 knitr::opts_knit$set(root.dir = normalizePath("/project/rohs_1070/analysis_data"))
```

```{r}
# Download Methylation Data
query <- GDCquery(project = "TCGA-SKCM",
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450",
                  data.format = "Dataframe")
GDCdownload(query)
methylation450 <- GDCprepare(query)

```

```{r}

# Retrieve, Process, and Store Methylation Data
methylation_clinical <- as.data.frame(methylation450@colData)
betas <- as.data.frame(methylation450@assays@data@listData)
cpg_sites <- as.data.frame(methylation450@rowRanges@elementMetadata)

column_mask <- ifelse(colnames(methylation_clinical) %in% c('treatments', 'primary_site', 'disease_type'), F, T)
methylation_clinical <- methylation_clinical[,column_mask]

# Write csv file 
write.csv(methylation_clinical, '/home1/kehuffma/490_cluster/r_review/skcm_methylation_clinical.csv')

site_mask <- !grepl('-', cpg_sites$gene) & !grepl(';', cpg_sites$gene) & !is.na(cpg_sites$gene) & complete.cases(betas)
betas <- betas[site_mask,]
cpg_sites <- cpg_sites[site_mask,]

write.csv(betas, '/home1/kehuffma/490_cluster/r_review/skcm_methylation_betas.csv')
write.csv(cpg_sites, '/home1/kehuffma/490_cluster/r_review/skcm_cpg_sites.csv')

```

```{r}
# Read in csv files
methylation_clinical <- read.csv('/home1/kehuffma/490_cluster/r_review/skcm_methylation_clinical.csv')
betas <- read.csv('/home1/kehuffma/490_cluster/r_review/skcm_methylation_betas.csv')
cpg_sites <- read.csv('/home1/kehuffma/490_cluster/r_review/skcm_cpg_sites.csv')
```

```{r}

# Masking definitions not in our interest
normal_tissue_mask <- methylation_clinical$definition == "Solid Tissue Normal"

methylation_clinical$tumor <- !normal_tissue_mask

mval <- t(apply(betas, 1, function(x) log2(x/(1-x))))

design <- model.matrix(~ tumor, data = methylation_clinical)
fit <- lmFit(mval, design)
fit2 <- eBayes(fit)
```

```{r}

# Data Processing for Visualization
dat <- data.frame(foldchange = fit[["coefficients"]][,2], logPvalue =  -log10(p.adjust(fit2[["p.value"]][,2],method='BY')), geneName = cpg_sites$gene)
dat$threshold <- as.factor(abs(dat$foldchange) < 1)

# Create Visualization
cols <- c("TRUE" = "grey", "FALSE" = "blue")
ggplot(data=dat, aes(x=foldchange, y = logPvalue, color=threshold)) +
  geom_point(alpha=.2, size=0.6) +
  scale_colour_manual(values = cols) +
  geom_vline(xintercept = 1, colour="#990000", linetype="dashed") + 
  geom_vline(xintercept = - 1, colour="#990000", linetype="dashed") +
  geom_hline(yintercept = 2, colour = "#990000", linetype="dashed") +
  theme(legend.position="none") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  theme_bw() +
  theme(legend.position = "none")

ggsave("/home1/kehuffma/490_cluster/r_review/Plot3.pdf")
```

# 4. Direct comparison of methylation status to transcriptional activity across non-metastatic vs metastatic patients:

```{r}
# Spotting genes of interest
downregulated <- res_df[(res_df$log2FoldChange < -2), 'gene_name']
hypermethylated <- dat[dat$foldchange > 2, 'geneName']
interest_genes <- intersect(downregulated, hypermethylated)
interest_genes
```

```{r}
# Direct Comparison of Transcriptional Activity to Methylation Status
GENE<-"MDS2"

gene_counts_mask <- skcm_genes$gene_name == GENE
gene_betas_mask <- cpg_sites$gene == GENE
skcm_clinical <- colData(skcm_se)[, ]

skcm_clinical_tumor <- skcm_clinical$definition == "Metastatic"
methylation_clinical_tumor <- methylation_clinical$definition == "Metastatic"

skcm_clinical_normal <- skcm_clinical$definition == "Primary solid Tumor"
methylation_clinical_normal <- methylation_clinical$definition == "Primary solid Tumor"

skcm_tumor <- as.numeric(skcm_counts[gene_counts_mask, skcm_clinical_tumor])
methylation_tumor <- (betas[gene_betas_mask, methylation_clinical_tumor])

skcm_normal <- as.numeric(skcm_counts[gene_counts_mask, skcm_clinical_normal])
methylation_normal <- (betas[gene_betas_mask, methylation_clinical_normal])
```

```{r}
# Create Boxplot
boxplot(skcm_normal, skcm_tumor, xlab='Group', ylab='Counts', names=c('Normal', 'Tumor'))

plot(1:dim(methylation_normal)[1], apply(methylation_normal, 1, mean), col='blue', type='h', lwd=3, ylab='Average Beta', xlab='CpG sites')
lines(1.3:(dim(methylation_normal)[1]+0.3), apply(methylation_tumor, 1, mean), col='red',
```

